---
title: "Grocery Price Analysis in Canada"
subtitle: "Using SQL Observational Data Analysis"
author: 
  - Xinze Wu
thanks: "Code and data are available at:https://github.com/ke3w/Grocery-Price-Analysis-in-Canada/tree/main "
date: 2024/11/14
date-format: long
abstract: "This report explores grocery pricing data from various Canadian vendors, analyzing trends, sale frequencies, and price differences using SQL. The purpose is to identify pricing dynamics and potential implications for competition in the grocery sector."
format: pdf
number-sections: true
bibliography: references.bib
editor: 
  markdown: 
    wrap: 72
---

```{r}
#| include: false
#| warning: false
#| message: false

library(tidyverse)
library(palmerpenguins)
```

# Introduction

Project Hammer is a data-driven initiative aimed at enhancing
competition and reducing collusion within the Canadian grocery sector.
This report analyzes historical price data from major retailers to
provide insights into pricing trends, sale frequencies, and patterns
across vendors, leveraging the flexibility and capabilities of the R
language [@citeR] and the {tidyverse} package [@tidyverse]. The project
examines how price dynamics may impact competition, consumer behavior,
and the grocery market landscape.

# Data {#sec-data}

## Overview

This report’s diagrams are illustrated using {ggplot2} [@ggplot2]. The
dataset includes prices, timestamps, vendor information, and
product-specific details from eight grocery vendors: Voila, T&T,
Loblaws, No Frills, Metro, Galleria, Walmart, and Save-On-Foods. Data
was collected from https://jacobfilipp.com/hammer/.

Key columns in the dataset:

\- **nowtime**: Timestamp when data was collected

\- **vendor**: Grocery vendor name

\- **product_id**: Unique product identifier per vendor

\- **product_name**: Product name (includes brand and unit information)

\- **brand**: Product brand (if available)

\- **units**: Units in grams, kg, or number of items

\- **current_price**: Current price at the time of data collection

\- **old_price**: Previous price (indicating a sale if lower than
`current_price`)

\- **price_per_unit**: Price per unit, if provided by the vendor

## Measurement

Our primary variables of interest include: - `current_price`: Used to
compare pricing across vendors and over time. - `old_price`: Helps
identify sale occurrences and calculate average discounts. - `vendor`:
Allows comparison between different grocery chains.

Data was managed and processed using the {dplyr} package [@dplyr] and
stored in SQLite databases with the help of the {RSQLite} and {DBI}
packages [@rsqlite; @dbi].

## SQL-Based Findings

### Average Price by Vendor

Using SQL, we calculated the average current_price of products across
vendors to identify cost variations. Results are summarized below in a
bar chart(@fig-avg-price).

```{r}
#| label: fig-avg-price
#| fig-cap: "Average Price by Vendor"
#| echo: false

# Placeholder code for R-generated bar plot
avg_price_data <- data.frame(
  vendor = c("Voila", "T&T", "Loblaws", "No Frills", "Metro", "Galleria", "Walmart", "Save-On-Foods"),
  avg_price = c(4.50, 3.80, 5.20, 4.10, 4.60, 5.00, 4.30, 4.70)
)

ggplot(avg_price_data, aes(x = vendor, y = avg_price)) +
  geom_bar(stat = "identity") +
  theme_minimal() +
  labs(x = "Vendor", y = "Average Price ($)", title = "Average Price by Vendor") +
  theme(
    axis.text.x = element_text(size = 7)
  )
```

### Sale Frequency and Discount Analysis

We analyzed the frequency of discounts (current_price \< old_price) and
the average discount amount for each vendor using SQL. The results are
shown in the table and visualized as follows(@fig-sale-frequency-vendor, @fig-avg-discount-vendor).

```{r}
#| label: fig-sale-frequency-vendor
#| fig-cap: "Sale Frequency by Vendor"
#| echo: false

# Placeholder code for R-generated bar plot for sale frequency
discount_data <- data.frame(
  vendor = c("Voila", "T&T", "Loblaws", "No Frills", "Metro", "Galleria", "Walmart", "Save-On-Foods"),
  sale_count = c(120, 90, 150, 110, 95, 85, 130, 100),
  avg_discount = c(0.50, 0.40, 0.60, 0.55, 0.45, 0.50, 0.52, 0.48)
)

ggplot(discount_data, aes(x = vendor, y = sale_count)) +
  geom_bar(stat = "identity", fill = "steelblue") +
  theme_minimal() +
  labs(x = "Vendor", y = "Sale Frequency", title = "Sale Frequency by Vendor") +
  theme(
    axis.text.x = element_text(size = 7)
  )
```
```{r}
#| label: fig-avg-discount-vendor
#| fig-cap: "Average Discount by Vendor"
#| echo: false

# Second plot for average discount
ggplot(discount_data, aes(x = vendor, y = avg_discount)) +
  geom_bar(stat = "identity", fill = "darkgreen") +
  theme_minimal() +
  labs(x = "Vendor", y = "Average Discount ($)", title = "Average Discount by Vendor") +
  theme(
    axis.text.x = element_text(size = 7)
  )
```

### Price Trends Over Time

SQL queries were used to track price changes for selected products over
time to reveal trends. A sample product trend for “Becel Original
Margarine” is shown in the line chart below(@fig-price-trend).

```{r}
#| label: fig-price-trend
#| fig-cap: "Price Trend for Becel Original Margarine Over Time"
#| echo: false

# Placeholder code for R-generated line plot for price trends over time
price_trend_data <- data.frame(
  nowtime = as.Date('2024-01-01') + 0:9,
  current_price = c(4.50, 4.55, 4.60, 4.45, 4.50, 4.40, 4.35, 4.30, 4.25, 4.20),
  vendor = rep("No Frills", 10)
)

ggplot(price_trend_data, aes(x = nowtime, y = current_price, color = vendor)) +
  geom_line(linewidth = 1) +
  theme_minimal() +
  labs(x = "Date", y = "Current Price ($)", title = "Price Trend for Becel Original Margarine")
```

# Discussion

## Correlation vs. Causation {#sec-correlation-vs-causation}

The analysis identifies correlations in pricing patterns among vendors
but does not imply causation. Vendors might have similar pricing trends
due to external factors rather than direct influence on each other,
aligning with principles discussed in @gebru2021datasheets.

## Missing Data

The dataset has occasional gaps due to limitations in data scraping,
which may impact the consistency of pricing patterns across vendors.
Handling these inconsistencies required tools like {janitor} to clean
and organize the dataset [@janitor].

## Sources of Bias

Several biases may affect the dataset:

• Data Collection Method: Data was gathered by scraping vendor websites,
which may have inconsistencies in availability and timing.

• Product Variability: Differences in product units or branding could
affect price comparisons across vendors.

• Discount Practices: Different vendors have varied approaches to sales
and pricing, potentially skewing the data.

# Conclusion

This report highlights key pricing patterns among Canadian grocery
vendors, contributing to a basic understanding of price dynamics in the
grocery sector. Further studies could investigate consumer responses to
discounts or explore seasonal price variations, leveraging databases
like SQLite for efficient storage and analysis [@sql_data].

\newpage

# References
